The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-10-09 paper.tex
Pequenos ajustes

\author{
Francisco Sant'Anna
\affil{Departamento de Inform\'atica, PUC--Rio}
Adriano Branco
\affil{Departamento de Inform\'atica, PUC--Rio}
Roberto Ierusalimschy
\affil{Departamento de Inform\'atica, PUC--Rio}
Noemi Rodriguez
\affil{Departamento de Inform\'atica, PUC--Rio}
Silvana Rossetto
\affil{Departamento de Ci\^encia da Computa\c{c}\~ao, UFRJ}
}

@@

\begin{abstract}
\CEU is a reactive language based on Esterel that targets constrained embedded 
platforms and ensures safe concurrency by handling threats at compile time, 
rather than at runtime.
%
Based on the synchronous programming model, our design allows for a simple 
reasoning about concurrency that enables compile-time analysis and results in 
deterministic and memory-safe programs.
%
% TODO: some uses?
%
In this work, we discuss the design of \CEU and propose a formal semantics 
focusing on its particular control mechanisms, such as parallel compositions, 
finalization, and stack-based internal events.
%
We also present an implementation with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

To illustrate threats related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, which does not compile in \CEU.
It describes the state machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input and output events represent the external interface of the protocol 
(ln. 1--2).
%
The protocol has to transmit a packet every minute (with \code{SEND\_ENQUEUE}), 
unless it receives a \code{RETRANSMIT} request to immediately re-transmit it, 
or a \code{STOP} request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (ln. 4);
the other periodically transmits the packet (ln. 6--19).
%
The periodic transmission is a loop that starts two other trails (ln. 7--18):
one handles the immediate retransmission request (ln. 8);
the other transmits the packet and waits for a confirmation (ln. 10--17).
%
%
The actual transmission (ln. 13--16) is enclosed with a \code{par/and} that 
takes at least one minute before looping (ln. 11), in accordance with the 
specification.
Note that the \code{emit SEND\_ENQUEUE} (ln. 15) is \emph{asynchronous}, 
handing to the radio driver a pointer to the lexically-scoped packet (ln. 13).
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SEND\_ACK} to acknowledge completion 
(ln. 16).
%
At any time, the client may request a retransmission (ln. 8), which terminates 
the \code{par/or} (ln. 7), aborts the ongoing transmission (ln. 15, if not 
idle), and restarts the loop (ln. 6).
%
The client may also request to stop the whole protocol at any time (ln. 4).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
\CEU tracks the interaction of \code{par/or} compositions with local variables 
and stateful output events and \C calls in order to preserve safe abortion of 
trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
accompany the output request.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

The communication between scripts in \CEU and the components in the \VM is 
mostly through events:
scripts \code{emit} requests through \code{output} events and \code{await} 
answers through \code{input} events.
%
Terra also provides system calls for initialization and configuration of 
components (e.g., \emph{getters} and \emph{setters}).
%
The left of Figure~\ref{lst.terra.defs} shows a \CEU interface with the 
available functionality for a customized \VM with temperature and radio 
components.
The right of the figure shows the associated binding for output events (ln.  
1--8), input events (ln. 10--14), and system calls (ln. 16--22).
%
Note that all applications for the customized \VM must comply with the same 
interface.
In contrast, the template-based \C back end allows applications to choose all 
possible combinations of functionalities from the underlying platform at 
compile time (as illustrated in Figure~\ref{lst.impl.tinyos}).

@@

Considering the compilation process illustrated in Figure~\ref{fig.impl} for 
the \C back end, the main difference resides in the \emph{Code Generation} 
phase, which here outputs assembly instructions for the Terra \VM, instead of 
\C.
%
The \VM includes special instructions to support complex and recurrent 
operations from the runtime of \CEU, such as dispatching events, clearing 
aborted trails
Basically, they map

@@

\begin{comment}
%-------------------------------------------------
\end{comment}

%-----------------------------------------------------
% Equivalent to the fig. 19

%------------------------------------------------------

\begin{comment}
Instead of generating code in C, Terra generates the vmt assembly code using 
the original \CEU labels for trails entry points.
In the final phase for code generation, Terra compiler generates the vmt 
bytecode file using the program address in place of \CEU labels.
The use of address allows the \CEU scheduler to execute directly a trail 
without a index table to replace the switch-case.
Related to memory layout and trail allocation, Terra uses the same data 
organization approach for all \CEU controls and variables.
All \CEU data and control structures for a specific program are determined at 
compile time.
Terra arranges all these structures in the beginning of virtual machine memory 
space.
The original \CEU run time control functions are maintained with a small 
modification where accesses to control structures have an off-set relative to 
the begin of the virtual machine memory space.
\end{comment}

@@

A number of virtual machines have been proposed for embedded systems.
%
The \emph{Sun SPOT} Java-based platform with the \emph{Squawk 
JVM}~\cite{wsn.sunspot} brings Java for the embedded domain, but requires a 
much powerful hardware.%
\footnote{The \emph{Sun SPOT} uses a 32-bit CPU with 4 Mbytes of FLASH and 512 
KBytes of SRAM.}
%
\emph{Darjeeling}~\cite{wsn.darjeeling} and \emph{TakaTuka}~\cite{wsn.takatuka} 
are complete \emph{JVMs} targeting constrained embedded systems with support 
for multithreading and garbage collection.
%
Java has antagonistic design choices in comparison to \CEU:
it does not impose static bounds on memory usage and execution time, and 
provides preemptive multithreading with synchronization primitives for 
accessing shared memory.
%
Plummer et al.~\cite{esterel.vm} propose a Esterel-based \VM with similar 
design choices to our work.
To reduce code size, the \VM has a instruction set to deal with event handling 
and concurrency constructs that are particular to Esterel.
However, the proposed \VM is only a proof of concept, with no support for 
arithmetic operations, external system calls, or remote reprogramming.

>>>>

\author{
Francisco Sant'Anna
\affil{Departamento de Inform\'atica, PUC--Rio}
Roberto Ierusalimschy
\affil{Departamento de Inform\'atica, PUC--Rio}
Noemi Rodriguez
\affil{Departamento de Inform\'atica, PUC--Rio}
Silvana Rossetto
\affil{Departamento de Ci\^encia da Computa\c{c}\~ao, UFRJ}
Adriano Branco
\affil{Departamento de Inform\'atica, PUC--Rio}
}

@@

\begin{abstract}
\CEU is a reactive language based on Esterel that targets constrained embedded 
platforms and ensures safe concurrency by handling threats at compile time.
%
Based on the synchronous programming model, our design allows for a simple 
reasoning about concurrency that enables compile-time analysis and results in 
deterministic and memory-safe programs.
%
% TODO: some uses?
%
We discuss the design of \CEU and propose a formal semantics for its particular 
control mechanisms, such as parallel compositions, finalization, and internal 
events.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

To illustrate threats related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, which does not compile in \CEU.
It describes the state machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input and output events represent the external interface of the protocol 
(ln. 1--2).
%
The protocol has to transmit a packet every minute (with \code{SEND\_ENQUEUE}), 
unless it receives a \code{RETRANSMIT} request to immediately re-transmit it, 
or a \code{STOP} request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (ln. 4);
the other periodically transmits the packet (ln. 6--19).
%
The periodic transmission is a loop that starts two other trails (ln. 7--18):
one handles the immediate retransmission request (ln. 8);
the other transmits the packet and waits for a confirmation (ln. 10--17).
%
%
The actual transmission (ln. 13--16) is enclosed with a \code{par/and} that 
takes at least one minute before looping (ln. 11), in accordance with the 
specification.
Note that the \code{emit SEND\_ENQUEUE} (ln. 15) is \emph{asynchronous}, 
handing to the radio driver a pointer to the lexically-scoped packet (ln. 13).
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SEND\_ACK} to acknowledge completion 
(ln. 16).
%
At any time, the client may request a retransmission (ln. 8), which terminates 
the \code{par/or} (ln. 7), aborts the ongoing transmission (ln. 15, if not 
idle), and restarts the loop (ln. 6).
%
The client may also request to stop the whole protocol at any time (ln. 4).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope (ln. 13), 
leaving behind a \emph{dangling pointer} in the radio driver, which will 
possibly transmit corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
\CEU tracks the interaction of \code{par/or} compositions with local variables 
and stateful output events calls in order to preserve safe abortion of trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
accompany the output request.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}, ln. 18) executes automatically on 
external abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

The communication between scripts in \CEU and the components in the \VM is 
mostly through events:
scripts \code{emit} requests through \code{output} events and \code{await} 
answers through \code{input} events.
%
Terra also provides system calls for initialization and configuration of 
components (e.g., \emph{getters} and \emph{setters}).
%
The left of Figure~\ref{lst.terra.defs} shows a \CEU interface with the 
available functionality for a customized \VM (with temperature and radio 
components).
The right of the figure shows the associated bindings for output events (ln.  
1--8), input events (ln. 10--14), and system calls (ln. 16--22).
%
Note that all applications for the customized \VM must comply with the same 
interface.
In contrast, the template-based \C back end (illustrated in 
Figure~\ref{lst.impl.tinyos}) allows applications to choose all possible 
combinations of functionalities from the underlying platform at compile time.

@@

Considering the compilation process illustrated in Figure~\ref{fig.impl} for 
the \C back end, the main difference resides in the \emph{Code Generation} 
phase, which here outputs assembly instructions for the Terra \VM, instead of 
statements in \C.
%
The \VM includes special instructions to support complex and recurrent 
operations from the runtime of \CEU, such as dispatching events, clearing 
aborted trails,
TODO
\begin{comment}
Basically, they map

@@

Instead of generating code in C, Terra generates the vmt assembly code using 
the original \CEU labels for trails entry points.
In the final phase for code generation, Terra compiler generates the vmt 
bytecode file using the program address in place of \CEU labels.
The use of address allows the \CEU scheduler to execute directly a trail 
without a index table to replace the switch-case.
Related to memory layout and trail allocation, Terra uses the same data 
organization approach for all \CEU controls and variables.
All \CEU data and control structures for a specific program are determined at 
compile time.
Terra arranges all these structures in the beginning of virtual machine memory 
space.
The original \CEU run time control functions are maintained with a small 
modification where accesses to control structures have an off-set relative to 
the begin of the virtual machine memory space.
\end{comment}

@@

A number of virtual machines have been proposed for embedded systems.
%
The \emph{Sun SPOT} Java-based platform with the \emph{Squawk JVM} brings Java 
for the embedded domain~\cite{wsn.sunspot}, but requires a much powerful 
hardware.%
\footnote{The \emph{Sun SPOT} uses a 32-bit CPU with 4 Mbytes of FLASH and 512 
KBytes of SRAM.}
%
\emph{Darjeeling}~\cite{wsn.darjeeling} and \emph{TakaTuka}~\cite{wsn.takatuka} 
are complete \emph{JVMs} targeting constrained embedded systems with support 
for multithreading and garbage collection.
%
Java has antagonistic design choices in comparison to \CEU:
it does not impose static bounds on memory usage and execution time, and 
provides preemptive multithreading with synchronization primitives for 
accessing shared memory.
%
Plummer et al.~\cite{esterel.vm} propose a Esterel-based \VM with similar 
design choices to our work.
To reduce code size, the \VM has a specialized instruction set to deal with 
events and concurrency constructs that are particular to Esterel.
However, the proposed \VM is only a proof of concept, with no support for 
arithmetic operations, external system calls, or remote reprogramming.
